\texttt{methyvim}: Targeted, Robust, and Model-free Analysis of Differential Methylation in R

### 2018-07-06 paper/paper_BiocF1000.Rmd
manuscript update

---
title: '`methyvim`: Targeted, Robust, and Model-free Analysis of Differential
  Methylation in R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing a general algorithm for the
  nonparametric estimation of treatment effects on DNA methylation at CpG sites
  throughout the genome, complete with straightforward statistical inference for
  such estimates. The approach leverages variable importance measures derived
  from statistical parameters arising in causal inference, defined in such a
  manner that they may be used to obtain targeted estimates of the relative
  importance of individual CpG sites with respect to a binary treatment assigned
  at the phenotype level. The procedure implemented is computationally
  efficient, incorporating a preliminary screening step to isolate a subset of
  sites for which there is cursory evidence of differential methylation as well
  as a unique multiple testing correction to control the False Discovery Rate
  with the same rigor as if all sites were tested. This technique for analysis
  of differentially methylated positions provides an avenue to incorporate
  flexible state-of-the-art machine learning algorithms into the estimation of
  differential methylation effects without the loss of interpretable statistical
  inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, machine learning, targeted minimum
  loss-based estimation
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

With the data structure above in hand, we let the estimand of interest be a
$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
through the _true_ (and unknown) probability distribution $P$. As a motivating
example, consider the case where $\Psi_j(P)$ is
\begin{equation}\label{vim_param}
  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
parameter of interest is a variant of the _average treatment effect_, which has
been the subject of much attention in statistical causal inference **CITATION
NEEDED**. As a measure of variable importance, we seek to estimate this target
parameter ($\Psi_j(P)$), which quantifies the effect of changing a treatment $A$
on the methylation $Y(j)$ of a CpG site $j$, accounting for any potential
confounding from phenotype-level covariates $W$ and observed methylation
$Y(S_j)$ at the neighboring sites $S_j$ of $j$. We propose a procedure that
estimates this target parameter $\Psi_j(P)$ across all CpG sites of interest
$j: 1, \ldots, J$. When data-adaptive regression procedures (i.e., machine
learning) are employed in estimating $\Psi_j(P)$, this produces a nonparametric
variable importance measure of differential methylation, allowing for the
identification of differentially methylated positions (DMPs) while avoiding many
assumptions common in the use of standard parametric regression procedures. We
propose estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which
allows for data-adaptive regression procedures to be employed in a
straightforward manner.

As a matter of practicality, we consider only estimating the target VIM
$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
span the full set of assayed CpG sites. In order to determine this subset of CpG
sites, we propose the use of a pre-screening procedure, which need be nothing
more than a method for differential methylation analysis that is computationally
less demanding than the method proposed here. Formally, let the pre-screening
procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
input a CpG site $j$ and returns a binary decision rule of whether to include
CpG site $j$ in a subset to be evaluated further or not. Details on the
implementation of pre-screening approaches is available in the sequel.

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
estimation, it is possible the a given CpG site $j$ may have _too many_
neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., Partitioning Around Medoids)
to select a _representative_ subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user [@kleinberg2003impossibility].

## Targeted Minimum Loss-Based Estimation and Inference

Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
minimum loss-based estimation (TMLE) to construct and evaluate an estimator
$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
target VIM is the average treatment effect, we make use of a TML estimator of
this parameter, which has been implemented for a general case in the
[`tmle`](https://CRAN.R-project.org/package=tmle) R package [@gruber2011tmle].
Generally speaking, a TML estimator is constructed from a few simple components:
(1) an estimator of the propensity score [@rosenbaum1983central] (often denoted
by $g(A, W)$), (2) an estimator of the outcome regression (often denoted $Q(A,
W)$), and (3) a targeting step applied to the initial estimator constructed from
the aforementioned components [@gruber2009targeted]. In the sequel, we describe
a few of the key mathematical properties of TML estimators relevant for
statistical inference, deferring extended discussion of the technical details to
@vdl2011targeted and @vdl2018targeted; moreover, as a matter of practicality, we
make use of the [`tmle`](https://CRAN.R-project.org/package=tmle)  R package of
@gruber2011tmle as a backend for computing estimators of relevant target
parameters in our software package, so the interested user is encouraged to
discuss the documentation of that software package for further details.

As noted previously, data-adaptive regression techniques may be used in the
construction of TML estimators. This is done by generating a stacked regression
model from a given set of machine learning algorithms via the Super Learner
algorithm [@wolpert1992stacked, breiman1996stacked, vdl2007super], which ensures
that the resulting ensemble learner satisifies important optimality properties
with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
@vdl2003unified]. Importantly, TML estimators are well suited for inference,
having an asymptotically normal limiting distribution
[@tsiatis2007semiparametric, @vdl2006targeted] $\sqrt{n}(\psi_{n, j} -
\Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)$, where $\sigma_{\text{EIF}}^2$ is
the variance of the efficient influence function of the target parameter
evaluated at the observed data. This allows for confidence intervals to be
constructed and and hypothesis tests (i.e., $H_{0, j}: \Psi_j(P) = 0$) to be
performed in a straightforward manner.

@@

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a _fully
nonparametric statistical model_. __The statistical algorithm consists in the
several major steps elaborated upon above and summarized below__:

@@

For the CpG sites that pass the pre-screening step, a user-specified target
parameter of interest is estimated independently at each site. _In all cases,
an estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation_.

For discrete-valued treatments or exposures:

@@

#### Future Work

@@

@@

>>>>

---
title: '`methyvim`: Targeted, Robust, and Model-free Analysis of Differential
  Methylation in R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing an algorithm for the
  nonparametric estimation of the effects of exposures on DNA methylation at CpG
  sites throughout the genome, complete with straightforward statistical
  inference for such estimates. The approach leverages variable importance
  measures derived from statistical parameters arising in causal inference,
  defined in such a manner that they may be used to obtain targeted estimates of
  the relative importance of individual CpG sites with respect to a binary
  treatment assigned at the phenotype level, thereby providing a new approach to
  identifying differentially methylated positions. The procedure implemented is
  computationally efficient, incorporating a preliminary screening step to
  isolate a subset of sites for which there is cursory evidence of differential
  methylation as well as a unique multiple testing correction to control the
  False Discovery Rate with the same rigor as would be available if all sites
  were subjected to testing. This novel technique for analysis of differentially
  methylated positions provides an avenue for incorporating flexible
  state-of-the-art data-adaptive regression procedures (i.e., machine learning)
  into the estimation of differential methylation effects without the loss of
  interpretable statistical inference for the estimated quantity.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, machine learning, targeted minimum
  loss-based estimation
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

@@

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a _fully
nonparametric statistical model_. The statistical algorithm consists in the
several major steps summarized below. Additional methodological details on the
use of targeted minimum loss-based estimation in this problem setting is
provided in a brief appendix (see section __Appendix__).

@@

For CpG sites that pass the pre-screening step, a user-specified target
parameter of interest is estimated independently at each site. _In all cases,
an estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation_.

Two popular target causal parameters for discrete-valued treatments or
exposures are

@@

<!--
#### Future Work

@@

-->

@@

# Appendix

With the data structure above in hand, we let the estimand of interest be a
$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
through the _true_ (and unknown) probability distribution $P$. As a motivating
example, consider the case where $\Psi_j(P)$ is
\begin{equation}\label{vim_param}
  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
parameter of interest is a variant of the _average treatment effect_, which has
been the subject of much attention in statistical causal inference
[@holland1986statistics, @hahn1998role, @hirano2003efficient]. As a measure of
variable importance, we seek to estimate this target parameter ($\Psi_j(P)$),
which quantifies the effect of changing a treatment $A$ on the methylation
$Y(j)$ of a CpG site $j$, accounting for any potential confounding from
phenotype-level covariates $W$ and observed methylation $Y(S_j)$ at the
neighboring sites $S_j$ of $j$. We propose a procedure that estimates this
target parameter $\Psi_j(P)$ across all CpG sites of interest $j: 1, \ldots, J$.
When data-adaptive regression procedures (i.e., machine learning) are employed
in estimating $\Psi_j(P)$, this produces a nonparametric variable importance
measure of differential methylation, allowing for the identification of
differentially methylated positions (DMPs) while avoiding many assumptions
common in the use of standard parametric regression procedures. We propose
estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which allows
for data-adaptive regression procedures to be employed in a straightforward
manner using the Super Learner algorithm for ensemble modeling [@vdl2007super,
@wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
@vdl2011targeted].

As a matter of practicality, we consider only estimating the target VIM
$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
span the full set of assayed CpG sites. In order to determine this subset of CpG
sites, we propose the use of a pre-screening procedure, which need be nothing
more than a method for differential methylation analysis that is computationally
less demanding than the method proposed here. Formally, let the pre-screening
procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
input a CpG site $j$ and returns a binary decision rule of whether to include
CpG site $j$ in a subset to be evaluated further or not. As an example, one
might consider employing a classical differential methylation analysis
procedure, such as the linear modeling approach of the
[`limma`](https://bioconductor.org/packages/limma) R package
[@smyth2004linear, @robinson2014statistical], using only CpG sites that pass a
reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
p-values) for the subsequent analysis steps in the `methyvim` pipeline. Details
on the implementation of pre-screening approaches is provided in the sequel.

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
estimation, it is possible the a given CpG site $j$ may have _too many_
neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., partitioning around medoids)
to select a _representative_ subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user [@kleinberg2003impossibility]. Note
that the goal of employing a clustering procedure in `methyvim` is to obtain a
smaller but still highly representative set of neighbors $S(j)$, so as to allow
for estimates of $\Psi_j(P)$ to account for as much confounding from neighboring
sites as is allowed by the available data.

## Targeted Minimum Loss-Based Estimation and Statistical Inference

Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
minimum loss-based estimation (TMLE) to construct and evaluate an estimator
$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
target VIM is based on the average treatment effect, we make use of a TML
estimator of this parameter, which has been implemented for a general case in
the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
[@gruber2011tmle]; users of `methyvim` may wish to consult the documentation of
that software package as a supplement. Generally speaking, a TML estimator is
constructed from a few simple components: (1) an estimator of the propensity
score [@rosenbaum1983central], often denoted by $g(A \mid W)$; (2) an estimator
of the outcome regression, often denoted $Q(A, W)$; and (3) a targeting step
applied to the initial estimators constructed from the aforementioned components
[@gruber2009targeted]. While we describe a few of the key properties of TML
estimators in the sequel, extended technical discussion is deferred to more
comprehensive work [@vdl2011targeted, @vdl2018targeted]. In order to construct
an estimate $\psi_{n, j}$ of $\Psi_j(P)$, it is necessary to accurately estimate
to nuisance parameters, these being the propensity score ($g(A \mid W)$) and the
outcome regression ($Q(A, W)$); moreover, data-adaptive regression procedures
may be used to obtain consistent estimates of these quantities through the
creation of a stacked regression model via the Super Learner algorithm
[@wolpert1992stacked, @breiman1996stacked, @vdl2007super], which ensures
that the resulting ensemble learner satisfies important optimality properties
with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
@vdl2003unified]. After constructing such estimates, an iterative procedure
(i.e., the targeting step) may be used to combine these individual estimates
into an estimate of the target parameter $\Psi_j(P)$.

Importantly, TML estimators are well-suited for statistical inference, having an
asymptotically normal limiting distribution [@tsiatis2007semiparametric,
@vdl2006targeted], which allows for a closed-form expression of the variance
to be derived:
\begin{equation}\label{eqn:lim_psi}
  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
\end{equation}
where $\sigma_{\text{EIF}}^2$ is the variance of the efficient influence
function, with respect to a nonparametric statistical model, of the target
parameter evaluated at the observed data. Such a convenience allows for
confidence intervals and hypothesis tests to be constructed (i.e.,
$H_{0, j}: \Psi_j(P) = 0$) in a straightforward manner.

### 2018-07-06 paper/paper_BiocF1000.tex
manuscript update

\begin{abstract}
We present \texttt{methyvim}, an R package implementing a general algorithm for the
nonparametric estimation of treatment effects on DNA methylation at CpG sites
throughout the genome, complete with straightforward statistical inference for
such estimates. The approach leverages variable importance measures derived
from statistical parameters arising in causal inference, defined in such a
manner that they may be used to obtain targeted estimates of the relative
importance of individual CpG sites with respect to a binary treatment assigned
at the phenotype level. The procedure implemented is computationally
efficient, incorporating a preliminary screening step to isolate a subset of
sites for which there is cursory evidence of differential methylation as well
as a unique multiple testing correction to control the False Discovery Rate
with the same rigor as if all sites were tested. This technique for analysis
of differentially methylated positions provides an avenue to incorporate
flexible state-of-the-art machine learning algorithms into the estimation of
differential methylation effects without the loss of interpretable statistical
inference.
\end{abstract}

@@

With the data structure above in hand, we let the estimand of interest be a
\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
example, consider the case where \(\Psi_j(P)\) is
\begin{equation}\label{vim_param}
  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where \(Y(S_j)\) is the subvector of \(Y\) that contains the measured methylation
values of the \emph{neighboring sites} of \(j\) (but not site \(j\) itself). This
parameter of interest is a variant of the \emph{average treatment effect}, which has
been the subject of much attention in statistical causal inference \textbf{CITATION
NEEDED}. As a measure of variable importance, we seek to estimate this target
parameter (\(\Psi_j(P)\)), which quantifies the effect of changing a treatment \(A\)
on the methylation \(Y(j)\) of a CpG site \(j\), accounting for any potential
confounding from phenotype-level covariates \(W\) and observed methylation
\(Y(S_j)\) at the neighboring sites \(S_j\) of \(j\). We propose a procedure that
estimates this target parameter \(\Psi_j(P)\) across all CpG sites of interest
\(j: 1, \ldots, J\). When data-adaptive regression procedures (i.e., machine
learning) are employed in estimating \(\Psi_j(P)\), this produces a nonparametric
variable importance measure of differential methylation, allowing for the
identification of differentially methylated positions (DMPs) while avoiding many
assumptions common in the use of standard parametric regression procedures. We
propose estimating \(\Psi_j(P)\) via targeted minimum loss-based estimation, which
allows for data-adaptive regression procedures to be employed in a
straightforward manner.

As a matter of practicality, we consider only estimating the target VIM
\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
span the full set of assayed CpG sites. In order to determine this subset of CpG
sites, we propose the use of a pre-screening procedure, which need be nothing
more than a method for differential methylation analysis that is computationally
less demanding than the method proposed here. Formally, let the pre-screening
procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
input a CpG site \(j\) and returns a binary decision rule of whether to include
CpG site \(j\) in a subset to be evaluated further or not. Details on the
implementation of pre-screening approaches is available in the sequel.

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., Partitioning Around Medoids)
to select a \emph{representative} subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user \citep{kleinberg2003impossibility}.

\hypertarget{targeted-minimum-loss-based-estimation-and-inference}{%
\subsection{Targeted Minimum Loss-Based Estimation and Inference}\label{targeted-minimum-loss-based-estimation-and-inference}}

Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
minimum loss-based estimation (TMLE) to construct and evaluate an estimator
\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
target VIM is the average treatment effect, we make use of a TML estimator of
this parameter, which has been implemented for a general case in the
\href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package \citep{gruber2011tmle}.
Generally speaking, a TML estimator is constructed from a few simple components:
(1) an estimator of the propensity score \citep{rosenbaum1983central} (often denoted
by \(g(A, W)\)), (2) an estimator of the outcome regression (often denoted \(Q(A, W)\)), and (3) a targeting step applied to the initial estimator constructed from
the aforementioned components \citep{gruber2009targeted}. In the sequel, we describe
a few of the key mathematical properties of TML estimators relevant for
statistical inference, deferring extended discussion of the technical details to
\citet{vdl2011targeted} and \citet{vdl2018targeted}; moreover, as a matter of practicality, we
make use of the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package of
\citet{gruber2011tmle} as a backend for computing estimators of relevant target
parameters in our software package, so the interested user is encouraged to
discuss the documentation of that software package for further details.

As noted previously, data-adaptive regression techniques may be used in the
construction of TML estimators. This is done by generating a stacked regression
model from a given set of machine learning algorithms via the Super Learner
algorithm \citep[breiman1996stacked, vdl2007super]{wolpert1992stacked}, which ensures
that the resulting ensemble learner satisifies important optimality properties
with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
\citet{vdl2003unified}]{vdl2004asymptotic}. Importantly, TML estimators are well suited for inference,
having an asymptotically normal limiting distribution
\citep[\citet{vdl2006targeted}]{tsiatis2007semiparametric} \(\sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)\), where \(\sigma_{\text{EIF}}^2\) is
the variance of the efficient influence function of the target parameter
evaluated at the observed data. This allows for confidence intervals to be
constructed and and hypothesis tests (i.e., \(H_{0, j}: \Psi_j(P) = 0\)) to be
performed in a straightforward manner.

@@

The core functionality of this package is made available via the eponymous
\texttt{methyvim} function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a \emph{fully
nonparametric statistical model}. \textbf{The statistical algorithm consists in the
several major steps elaborated upon above and summarized below}:

@@

For the CpG sites that pass the pre-screening step, a user-specified target
parameter of interest is estimated independently at each site. \emph{In all cases,
an estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation}.

For discrete-valued treatments or exposures:

@@

\hypertarget{future-work}{%
\paragraph{Future Work}\label{future-work}}

Support for continuous-valued treatments or exposures is \emph{planned and ongoing}
but not yet available in stable releases of the \texttt{methyvim} software package.
Future releases will allow users to assess continuous-valued treatments by
relying on parameters estimable through implementations available in the
following software packages:

\begin{itemize}
\item
  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}
  (R package \texttt{tmle.npvi}). The effect of a continuous-valued exposure or
  treatment (the observed methylation at a target CpG site) on an outcome of
  interest is estimated, controlling for the observed methylation value at all
  other CpG sites in the same neighborhood as the target site. This uses a
  parameter that compares values of the treatment against a user-specified
  reference value taken to be the null value. In particular, the implementation
  to be provided is designed to assess the effect of differential methylation at
  the target CpG site on an outcome of interest (e.g., survival), providing a
  nonparametric evaluation of the impact of methylation at the target site.
\item
  The causal effect of shifting the value of an observed intervention, defined
  as the counterfactual outcome under a posited shift of a continuous-value
  treatment of interest using stochastic intervention policies
  \citep[\citet{diaz2018stochastic}, \citet{hejazi2018txshift}]{munoz2012population} (R package
  \texttt{txshift}). The value of an outcome of interest under an unobserved value of
  the (continuous-valued) treatment, specified through a user-provided
  \emph{additive shift} of the observed treatment, may be data-adaptively estimated.
  This allows for the effect of changes/shifts in the observed methylation at a
  CpG site on an outcome of interest to be evaluated, providing a nonparametric
  evalution of the relative importance of a particular target CpG site with
  respect to another variable measured in the same study.
\end{itemize}

@@

>>>>

\begin{abstract}
We present \texttt{methyvim}, an R package implementing an algorithm for the
nonparametric estimation of the effects of exposures on DNA methylation at CpG
sites throughout the genome, complete with straightforward statistical
inference for such estimates. The approach leverages variable importance
measures derived from statistical parameters arising in causal inference,
defined in such a manner that they may be used to obtain targeted estimates of
the relative importance of individual CpG sites with respect to a binary
treatment assigned at the phenotype level, thereby providing a new approach to
identifying differentially methylated positions. The procedure implemented is
computationally efficient, incorporating a preliminary screening step to
isolate a subset of sites for which there is cursory evidence of differential
methylation as well as a unique multiple testing correction to control the
False Discovery Rate with the same rigor as would be available if all sites
were subjected to testing. This novel technique for analysis of differentially
methylated positions provides an avenue for incorporating flexible
state-of-the-art data-adaptive regression procedures (i.e., machine learning)
into the estimation of differential methylation effects without the loss of
interpretable statistical inference for the estimated quantity.
\end{abstract}

@@

@@

The core functionality of this package is made available via the eponymous
\texttt{methyvim} function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a \emph{fully
nonparametric statistical model}. The statistical algorithm consists in the
several major steps summarized below. Additional methodological details on the
use of targeted minimum loss-based estimation in this problem setting is
provided in a brief appendix (see section \textbf{Appendix}).

@@

For CpG sites that pass the pre-screening step, a user-specified target
parameter of interest is estimated independently at each site. \emph{In all cases,
an estimator of the parameter of interest is constructed via targeted minimum
loss-based estimation}.

Two popular target causal parameters for discrete-valued treatments or
exposures are

@@

@@

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

With the data structure above in hand, we let the estimand of interest be a
\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
example, consider the case where \(\Psi_j(P)\) is
\begin{equation}\label{vim_param}
  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
\end{equation}
where \(Y(S_j)\) is the subvector of \(Y\) that contains the measured methylation
values of the \emph{neighboring sites} of \(j\) (but not site \(j\) itself). This
parameter of interest is a variant of the \emph{average treatment effect}, which has
been the subject of much attention in statistical causal inference
\citep[\citet{hahn1998role}, \citet{hirano2003efficient}]{holland1986statistics}. As a measure of
variable importance, we seek to estimate this target parameter (\(\Psi_j(P)\)),
which quantifies the effect of changing a treatment \(A\) on the methylation
\(Y(j)\) of a CpG site \(j\), accounting for any potential confounding from
phenotype-level covariates \(W\) and observed methylation \(Y(S_j)\) at the
neighboring sites \(S_j\) of \(j\). We propose a procedure that estimates this
target parameter \(\Psi_j(P)\) across all CpG sites of interest \(j: 1, \ldots, J\).
When data-adaptive regression procedures (i.e., machine learning) are employed
in estimating \(\Psi_j(P)\), this produces a nonparametric variable importance
measure of differential methylation, allowing for the identification of
differentially methylated positions (DMPs) while avoiding many assumptions
common in the use of standard parametric regression procedures. We propose
estimating \(\Psi_j(P)\) via targeted minimum loss-based estimation, which allows
for data-adaptive regression procedures to be employed in a straightforward
manner using the Super Learner algorithm for ensemble modeling \citep[
\citet{wolpert1992stacked}, \citet{breiman1996stacked}, \citet{gruber2009targeted},
\citet{vdl2011targeted}]{vdl2007super}.

As a matter of practicality, we consider only estimating the target VIM
\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
span the full set of assayed CpG sites. In order to determine this subset of CpG
sites, we propose the use of a pre-screening procedure, which need be nothing
more than a method for differential methylation analysis that is computationally
less demanding than the method proposed here. Formally, let the pre-screening
procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
input a CpG site \(j\) and returns a binary decision rule of whether to include
CpG site \(j\) in a subset to be evaluated further or not. As an example, one
might consider employing a classical differential methylation analysis
procedure, such as the linear modeling approach of the
\href{https://bioconductor.org/packages/limma}{\texttt{limma}} R package
\citep[\citet{robinson2014statistical}]{smyth2004linear}, using only CpG sites that pass a
reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
p-values) for the subsequent analysis steps in the \texttt{methyvim} pipeline. Details
on the implementation of pre-screening approaches is provided in the sequel.

Due to the data-adaptive nature of the regression procedures employed in
evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
the inclusion of too many neighbors when controlling for potential confounders
may lead to instability in the estimates produced. In such cases, we propose and
implement the use of a clustering technique (e.g., partitioning around medoids)
to select a \emph{representative} subset of neighbors. Formally, there is likely no
best choice of a specific clustering algorithm, so we leave this aspect of the
proposed algorithm as flexible for the user \citep{kleinberg2003impossibility}. Note
that the goal of employing a clustering procedure in \texttt{methyvim} is to obtain a
smaller but still highly representative set of neighbors \(S(j)\), so as to allow
for estimates of \(\Psi_j(P)\) to account for as much confounding from neighboring
sites as is allowed by the available data.

\hypertarget{targeted-minimum-loss-based-estimation-and-statistical-inference}{%
\subsection{Targeted Minimum Loss-Based Estimation and Statistical Inference}\label{targeted-minimum-loss-based-estimation-and-statistical-inference}}

Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
minimum loss-based estimation (TMLE) to construct and evaluate an estimator
\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
target VIM is based on the average treatment effect, we make use of a TML
estimator of this parameter, which has been implemented for a general case in
the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package
\citep{gruber2011tmle}; users of \texttt{methyvim} may wish to consult the documentation of
that software package as a supplement. Generally speaking, a TML estimator is
constructed from a few simple components: (1) an estimator of the propensity
score \citep{rosenbaum1983central}, often denoted by \(g(A \mid W)\); (2) an estimator
of the outcome regression, often denoted \(Q(A, W)\); and (3) a targeting step
applied to the initial estimators constructed from the aforementioned components
\citep{gruber2009targeted}. While we describe a few of the key properties of TML
estimators in the sequel, extended technical discussion is deferred to more
comprehensive work \citep[\citet{vdl2018targeted}]{vdl2011targeted}. In order to construct
an estimate \(\psi_{n, j}\) of \(\Psi_j(P)\), it is necessary to accurately estimate
to nuisance parameters, these being the propensity score (\(g(A \mid W)\)) and the
outcome regression (\(Q(A, W)\)); moreover, data-adaptive regression procedures
may be used to obtain consistent estimates of these quantities through the
creation of a stacked regression model via the Super Learner algorithm
\citep[\citet{breiman1996stacked}, \citet{vdl2007super}]{wolpert1992stacked}, which ensures
that the resulting ensemble learner satisfies important optimality properties
with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
\citet{vdl2003unified}]{vdl2004asymptotic}. After constructing such estimates, an iterative procedure
(i.e., the targeting step) may be used to combine these individual estimates
into an estimate of the target parameter \(\Psi_j(P)\).

Importantly, TML estimators are well-suited for statistical inference, having an
asymptotically normal limiting distribution \citep[
\citet{vdl2006targeted}]{tsiatis2007semiparametric}, which allows for a closed-form expression of the variance
to be derived:
\begin{equation}\label{eqn:lim_psi}
  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
\end{equation}
where \(\sigma_{\text{EIF}}^2\) is the variance of the efficient influence
function, with respect to a nonparametric statistical model, of the target
parameter evaluated at the observed data. Such a convenience allows for
confidence intervals and hypothesis tests to be constructed (i.e.,
\(H_{0, j}: \Psi_j(P) = 0\)) in a straightforward manner.
